15.071 - Analytics Edge
------------------------------------------------------------------
#### The Framingham Heart Study
#### Tarek Dib
#### Date: March 23, 2014

## *Introduction*
The model's main purpose is to predict the 10-year risk of Coronary Heart Disease (CHD). CHD is a disease of the blood vessels supplying the heart. Heart disease has been the leading cause of death worldwide since 1921. In 2008, 7.3 million people died from CHD. The model is built using the following proceedure:
    Identify risk factors
    Collect data
    Predict heart disease
    Collect more data
    Validate model
    Define interventions using model 

### *Independent variables:*
#### *Demographic risk factors*
    male: sex of patient
    age: age in years at first examination
    education: Some high school (1), high school/GED (2), some college/vocational school (3), college (4)
#### *Behavioral risk factors*
    currentSmoker
    cigsPerDay: Smoking behavior 
#### *Medical history risk factors*
    BPmeds: On blood pressure medication at time of first examination
    prevalentStroke: Previously had a stroke
    prevalentHyp: Currently hypertensive
    diabetes: Currently has diabetes 
#### *Risk factors from first examination*
    totChol: Total cholesterol (mg/dL)
    sysBP: Systolic blood pressure
    diaBP: Diastolic blood pressure
    BMI: Body Mass Index, weight (kg)/height (m)^2
    heartRate: Heart rate (beats/minute)
    glucose: Blood glucose level (mg/dL)
More information about the data set used in the study can be found at: https://biolincc.nhlbi.nih.gov/static/studies/teaching/framdoc.pdf
## *Summary and Classification*
```{r}
# Read in the dataset
setwd("Analytics/Weeks/Week3/Data")
framingham <- read.csv("framingham.csv")

# Structure of data
str(framingham)

# Load the library caTools
library(caTools)

# Randomly split the data into training and testing sets
set.seed(1000)
# One needs to put between 50% and 80% of data in the training set
split = sample.split(framingham$TenYearCHD, SplitRatio = 0.65)

# Split up the data using subset
train = subset(framingham, split==TRUE)
test = subset(framingham, split==FALSE)
```

## *The Model, Confusion Matrix and Model Accuracy*
```{r}
# Logistic Regression Model
framinghamLog = glm(TenYearCHD ~ ., data = train, family=binomial)
summary(framinghamLog)

# Predictions on the test set
predictTest = predict(framinghamLog, type="response", newdata=test)

# Confusion matrix with threshold of 0.5
ConfMat <- table(test$TenYearCHD, predictTest > 0.5)
ConfMat

# Accuracy
(ConfMat[1,1]+ConfMat[2,2])/(ConfMat[1,1]+ConfMat[1,2]+ConfMat[2,1]+ConfMat[2,2])

# Baseline accuracy
(1069+6)/(1069+6+187+11) 
```

## *Test AUC*
```{r}
library(ROCR)
ROCRpred = prediction(predictTest, test$TenYearCHD)
as.numeric(performance(ROCRpred, "auc")@y.values)
```
