Climate Change, Analytics Edge, MITx: 15.073x
========================================================
### Tarek Dib

## *Intoduction*
There have been many studies documenting that the average global temperature has been increasing over the last century. The consequences of a continued rise in global temperature will be dire. Rising sea levels and an increased frequency of extreme weather events will affect billions of people.

We will study the relationship between average global temperature and several other factors. Climate data are from May 1983 to December 2008. The available variables include:

    Year: the observation year.
    Month: the observation month.
    Temp: the difference in degrees Celsius between the average global temperature in that period and a reference value. This data comes from the Climatic Research Unit at the University of East Anglia.
    CO2, N2O, CH4, CFC.11, CFC.12: atmospheric concentrations of carbon dioxide (CO2), nitrous oxide (N2O), methane  (CH4), trichlorofluoromethane (CCl3F; commonly referred to as CFC-11) and dichlorodifluoromethane (CCl2F2; commonly referred to as CFC-12), respectively. This data comes from the ESRL/NOAA Global Monitoring Division.
        CO2, N2O and CH4 are expressed in ppmv (parts per million by volume  -- i.e., 397 ppmv of CO2 means that CO2 constitutes 397 millionths of the total volume of the atmosphere)
        CFC.11 and CFC.12 are expressed in ppbv (parts per billion by volume). 
    Aerosols: the mean stratospheric aerosol optical depth at 550 nm. This variable is linked to volcanoes, as volcanic eruptions result in new particles being added to the atmosphere, which affect how much of the sun's energy is reflected back into space. This data is from the Godard Institute for Space Studies at NASA.
    TSI: the total solar irradiance (TSI) in W/m2 (the rate at which the sun's energy is deposited per unit area). Due to sunspots and other solar phenomena, the amount of energy that is given off by the sun varies substantially with time. This data is from the SOLARIS-HEPPA project website.
    MEI: multivariate El Nino Southern Oscillation index (MEI), a measure of the strength of the El Nino/La Nina-Southern Oscillation (a weather effect in the Pacific Ocean that affects global temperatures). This data comes from the ESRL/NOAA Physical Sciences Division.


## *Creating Our First Model*
First, we will split the data into training and test sets. Training set includes data prior to 2006, and the test set includes 2007 and 2008 data.
```{r}
setwd("Analytics/Weeks/Week2/Data")
climate <- read.csv("climate_change.csv")
trn <- subset(climate, climate$Year <= 2006)
tst <- subset(climate, climate$Year > 2006)
# Build model with all variables as predictors except Year and Month. Response is Temp
model1 <- lm(Temp~.-Year - Month, data=trn)
summary(model1)
# Correlation matrix of the training set
cor(trn)
# Remove the highly correlated independent variables from the above mode. These variables accordingto the correlation matrix are: CO2, CH4, CFC.12, CFC.11. Thus build a model with the following variables: MEI, TSI, Aerosols, N2O
model2 <- lm(Temp ~ MEI + TSI + Aerosols + N2O, data=trn)
summary(model2)
# Model selection using the step function
model3 <- step(model1)
# Note: It is interesting to note that the step function does not address the collinearity of the variables, except that adding highly correlated variables will not improve the R2 significantly. The consequence of this is that the step function will not necessarily produce a very interpretable model - just a model that has balanced quality and simplicity for a particular weighting of quality and simplicity (AIC).

# Use the test data set to predict temperature difference
temp.predicted <- predict(model3, newdata=tst)
summary(temp.predicted)
```




```{r fig.width=12, fig.height=6}
plot(statedata$Income, statedata$Life.Exp)
```

