MITx: 15.071x The Analytics Edge - MARKET SEGMENTATION FOR AIRLINES
========================================================
# Name: Tarek Dib
# Date: 04/21/2014

## *Introduction*
Market segmentation is a strategy that divides a broad target market of customers into smaller, more similar groups, and then designs a marketing strategy specifically for each group. Clustering is a common technique for market segmentation since it automatically finds similar groups given a data set. 

In this problem, we'll see how clustering can be used to find similar groups of customers who belong to an airline's frequent flyer program. The airline is trying to learn more about its customers so that it can target different customer segments with different types of mileage offers. 

The file AirlinesCluster.csv contains information on 3,999 members of the frequent flyer program. This data comes from the textbook "Data Mining for Business Intelligence," by Galit Shmueli, Nitin R. Patel, and Peter C. Bruce.

There are seven different variables in the dataset, described below:

    Balance = number of miles eligible for award travel
    QualMiles = number of miles qualifying for TopFlight status
    BonusMiles = number of miles earned from non-flight bonus transactions in the past 12 months
    BonusTrans = number of non-flight bonus transactions in the past 12 months
    FlightMiles = number of flight miles in the past 12 months
    FlightTrans = number of flight transactions in the past 12 months
    DaysSinceEnroll = number of days since enrolled in the frequent flyer program

## *NORMALIZING THE DATA*
```{r}
# Data
setwd("/home/tarek/Analytics/Weeks/Week6-Clustering/Data")
airlines <- read.csv("AirlinesCluster.csv")
summary(airlines)

# Create a normalized data frame called "airlinesNorm".If we don't normalize the data, the variables that are on a larger scale will contribute much more to the distance calculation, and thus will dominate the clustering.

library(caret)
# Normalize the variables in the airlines data frame by using the preProcess function in the "caret" package
# Pre process
preproc = preProcess(airlines)
# Normalize
airlinesNorm = predict(preproc, airlines)
summary(airlinesNorm)
```

## *HIERARCHICAL CLUSTERING*
```{r}
airlinesDist <- dist(airlinesNorm, method="euclidean")
airlinesHierClust <- hclust(airlinesDist, method="ward") 
plot(airlinesHierClust)

# Pick 5 clusters
clusterGroups = cutree(airlinesHierClust, k = 5)
HierCluster = split(airlinesNorm, clusterGroups)
# Number of data points in cluster 1
nrow(HierCluster[[1]])

balanceAvg <- tapply(airlines$Balance, clusterGroups, mean)
qualMilesAvg <- tapply(airlines$QualMiles, clusterGroups, mean)
bonusMilesAvg <- tapply(airlines$BonusMiles, clusterGroups, mean)
bonusTransAvg <- tapply(airlines$BonusTrans, clusterGroups, mean)
flightMilesAvg <- tapply(airlines$FlightMiles, clusterGroups, mean)
flightTransAvg <- tapply(airlines$FlightTrans, clusterGroups, mean)
daysAvg <- tapply(airlines$DaysSinceEnroll, clusterGroups, mean)
df <- as.data.frame(cbind(balanceAvg, qualMilesAvg, bonusMilesAvg, bonusTransAvg, flightMilesAvg, flightTransAvg, daysAvg))
which.max(df$balanceAvg); 
which.max(df$qualMilesAvg);
which.max(df$bonusMilesAvg); 
which.max(df$bonusTransAvg); 
which.max(df$flightMilesAvg); 
which.max(df$flightTransAvg);
which.max(df$daysAvg)
```

## *K-means Clustering*
```{r}
k = 5
set.seed(88)
KmeansCluster <- kmeans(airlinesNorm, centers = k, iter.max=1000)
airlinesClusters <- KmeansCluster$cluster
table(airlinesClusters) > 1000

KmeansCluster$centers
