MITx: 15.071x The Analytics Edge - PREDICTING MEDICAL COSTS WITH CLUSTER-THEN-PREDICT
========================================================
# Name: Tarek Dib
# Date: 04/22/2014

## *Introduction*
In the second lecture sequence this week, we heard about cluster-then-predict, a methodology in which you first cluster observations and then build cluster-specific prediction models. In the lecture sequence, we saw how this methodology helped improve the prediction of heart attack risk. In this assignment, we'll use cluster-then-predict to predict future medical costs using medical claims data.

In Week 4, we discussed the importance of high-quality predictions of future medical costs based on information available in medical claims data. In this problem, you will predict future medical claims using part of the DE-SynPUF dataset, published by the United States Centers for Medicare and Medicaid Services (CMS). This dataset, available in reimbursement.csv, is structured to represent a sample of patients in the Medicare program, which provides health insurance to Americans aged 65 and older as well as some younger people with certain medical conditions. To protect the privacy of patients represented in this publicly available dataset, CMS performs a number of steps to anonymize the data, so we would need to re-train the models we develop in this problem on de-anonymized data if we wanted to apply our models in the real world.

The observations in the dataset represent a 1% random sample of Medicare beneficiaries in 2008, limited to those still alive at the end of 2008. The dependent variable, reimbursement2009, represents the total value of all Medicare reimbursements for a patient in 2009, which is the cost of the patient's care to the Medicare system. The following independent variables are available:

    age: The patient's age in years at the beginning of 2009
    alzheimers: Binary variable for whether the patient had diagnosis codes for Alzheimer's       disease or a related disorder in 2008
    arthritis: Binary variable for whether the patient had diagnosis codes for rheumatoid arthritis or osteoarthritis in 2008
    cancer: Binary variable for whether the patient had diagnosis codes for cancer in 2008
    copd: Binary variable for whether the patient had diagnosis codes for Chronic Obstructive Pulmonary Disease (COPD) in 2008
    depression: Binary variable for whether the patient had diagnosis codes for depression in 2008
    diabetes: Binary variable for whether the patient had diagnosis codes for diabetes in 2008
    heart.failure: Binary variable for whether the patient had diagnosis codes for heart failure in 2008
    ihd: Binary variable for whether the patient had diagnosis codes for ischemic heart disease (IHD) in 2008
    kidney: Binary variable for whether the patient had diagnosis codes for chronic kidney disease in 2008
    osteoporosis: Binary variable for whether the patient had diagnosis codes for osteoporosis in 2008
    stroke: Binary variable for whether the patient had diagnosis codes for a stroke/transient ischemic attack (TIA) in 2008
    reimbursement2008: The total amount of Medicare reimbursements for this patient for 2008

## *PREPARING THE DATASET*
```{r}
# Data
setwd("/home/tarek/Analytics/Weeks/Week6-Clustering/Data")
claims <- read.csv("reimbursement.csv")

# Subset of claims. Patients who had at least one of the chronic conditions
has.condition = subset(claims, alzheimers == 1 | arthritis == 1 | cancer == 1 | copd == 1 | depression == 1 | diabetes == 1 | heart.failure == 1 | ihd == 1 | kidney == 1 | osteoporosis == 1 | stroke == 1)
# Ratio
nrow(has.condition) / nrow(claims)
# Maximum correlation between independent variables
sort(cor(claims[-(13:14)]))[132]

# Distribution of reimbursement in 2008
hist(claims$reimbursement2008)

# Transform the dependent variables. +1 so that we don't get log(0) which is -infinity
claims$reimbursement2008 = log(claims$reimbursement2008+1)
claims$reimbursement2009 = log(claims$reimbursement2009+1)

hist(claims$reimbursement2009, main="Log of Reimbursement in 2009")
# About 20% of beneficiaries had $0 reimbursement in 2009
sum(claims$reimbursement2009==0) / length(claims$reimbursement2009)
```

## *INITIAL LINEAR REGRESSION MODEL*
```{r}
set.seed(144)
spl = sample(1:nrow(claims), size=0.7*nrow(claims))
train = claims[spl,]
test = claims[-spl,]

lm.claims <- lm(reimbursement2009~., data=train)
summary(lm.claims)

# MSE
predTest <- predict(lm.claims, newdata=test)
SSE <- sum((predTest - test$reimbursement2009)^2)
rms.lm <- sqrt(SSE/nrow(test))
rms.lm

# The naive baseline predicts the average of the dependent variable (reimbursement2009) on the training set! baseline model is mean(train$reimbursement2009). The testing MSE is:
rms.baseline <- sqrt(mean((mean(train$reimbursement2009) - test$reimbursement2009)^2))
rms.baseline

# Smart baseline model: predict that a patient's medical costs would be equal to their costs in the previous year
smartModel <- train$reimbursement2008
# RMSE
sqrt(mean((test$reimbursement2008-test$reimbursement2009)^2))
```

## *CLUSTERING MEDICARE BENEFICIARIES*
```{r}
# remove the dependent variable using the following commands:
train.limited = train
train.limited$reimbursement2009 = NULL
test.limited = test
test.limited$reimbursement2009 = NULL

# In cluster-then-predict, our final goal is to predict the dependent variable, which is unknown to us at the time of prediction. Therefore, if we need to know the outcome value to perform the clustering, the methodology is no longer useful for prediction of an unknown outcome value.

#This is an important point that is sometimes mistakenly overlooked. If you use the outcome value to cluster, you might conclude your method strongly outperforms a non-clustering alternative. However, this is because it is using the outcome to determine the clusters, which is not valid.

# Normalize
library(caret)
preproc = preProcess(train.limited)
train.norm = predict(preproc, train.limited)
test.norm = predict(preproc, test.limited)
mean(train.norm$arthritis)
```

## *K-means Clustering*
```{r}
k = 3
set.seed(144)
km <- kmeans(train.norm, centers = k)
kmClust <- km$cluster
table(kmClust)

km$centers

# Use the flexclust package to obtain training set and testing set cluster assignments for our observations
library(flexclust)
km.kcca = as.kcca(km, train.norm)
cluster.train = predict(km.kcca)
cluster.test = predict(km.kcca, newdata=test.norm)
sum(cluster.test == 2)
```

## *CLUSTER-SPECIFIC PREDICTIONS*
```{r}
train1 <- subset(train, cluster.train==1)
train2 <- subset(train, cluster.train==2)
train3 <- subset(train, cluster.train==3)
mean(train1$reimbursement2009); mean(train2$reimbursement2009); mean(train3$reimbursement2009)

test1 = subset(test, cluster.test == 1)
test2 = subset(test, cluster.test == 2)
test3 = subset(test, cluster.test == 3)

lm1 <- lm(reimbursement2009~., data=train1)
lm2 <- lm(reimbursement2009~., data=train2)
lm3 <- lm(reimbursement2009~., data=train3)
summary(lm1); summary(lm2); summary(lm3)

pred.Test1 <- predict(lm1, newdata=test1)
pred.Test2 <- predict(lm2, newdata=test2)
pred.Test3 <- predict(lm3, newdata=test3)

pred.TestMeans <- c(mean(pred.Test1), mean(pred.Test2), mean(pred.Test3))
names(pred.TestMeans) = c("pred.Test1", "pred.Test2", "pred.Test3")
# Which vector of test-set predictions has the smallest average predicted reimbursement amount?
which.min(pred.TestMeans)

# Obtain the test-set RMSE for each cluster
RMSE.Test1 <- sqrt(mean((pred.Test1 - test1$reimbursement2009)^2))
RMSE.Test2 <- sqrt(mean((pred.Test2 - test2$reimbursement2009)^2))
RMSE.Test3 <- sqrt(mean((pred.Test3 - test3$reimbursement2009)^2))
RMSE.Test <- c(RMSE.Test1, RMSE.Test2, RMSE.Test3)
names(RMSE.Test) = c(RMSE.Test1, RMSE.Test2, RMSE.Test3)
# Which cluster has the largest test-set RMSE?
which.max(RMSE.Test)

# To compute the overall test-set RMSE of the cluster-then-predict approach, we can combine all the test-set predictions into a single vector and all the true outcomes into a single vector.
all.predictions = c(pred.Test1, pred.Test2, pred.Test3)
all.outcomes = c(test1$reimbursement2009, test2$reimbursement2009, test3$reimbursement2009)
# What is the test-set RMSE of the cluster-then-predict approach
sqrt(mean((all.predictions - all.outcomes)^2))
```

We see a modest improvement over the original linear regression model, which is typical in situations where the observations do not cluster strongly into different "types" of observations. However, it is often a good idea to try the cluster-then-predict approach on datasets with a large number of observations to see if you can improve the accuracy of your model.